# PalEON Summer Course: MCMC with JAGS module
## August 2014
## Chris Paciorek


In this module, we'll do some hands-on work to fit a time series model via MCMC with JAGS. 


Let's work through our mixed model example and the DLM example using
JAGS (or BUGS).


\subsection{Mixed model}

Here we'll use simulated data.

```{r mixedModelData}
theta <- 2
sig <- 2
nMoms <- 10

## IGNORE THE MAN BEHIND THE CURTAIN ####
set.seed(0)
mus <- rgamma(nMoms, shape = 2, scale = 1)
nPups <- rbinom(nMoms, 15, prob = .4) 
momIds <- rep(1:nMoms, times = nPups)
y <- rnorm(sum(nPups), mus[momIds], sig)
n <- sum(nPups)
## IGNORE THE CODE ABOVE
```

Here's the BUGS code for the model.

```{r bugsCode_mixedModel}
### model function (note this looks like R code but is really JAGS code for defining a model
mixedModel <- function(){
  # (hyper)priors
  sig ~ dunif(0, 100)
  tau ~ dunif(0, 100)
  theta ~ dnorm(0,.00001)
  # some transformations to get precisions from sd's
  tau2Inv <- 1/(tau^2)
  sig2Inv <- 1/(sig^2)
  # latent value layer
  for(i in 1:nMoms){
    mus[i] ~ dnorm(theta, tau2Inv)
  }
  # likelihood
  for(j in 1:n)
    {
      y[j] ~ dnorm(mus[momIds[j]], sig2Inv)
    }
}
```

Here's how we run it from R.

```{r run_mixed}
require(R2jags, quietly = TRUE) 
out <- jags(data = list(y = y, n = n, nMoms = nMoms, momIds = momIds),
  parameters.to.save = c('theta','sig','tau', 'mus'), n.chains = 1,
  n.iter = 2000, n.burnin = 1000, model.file = mixedModel, DIC = FALSE)
```


A few things to check to see if the MCMC is performing well:
- Look at the trace plots and see if the initial part of the chain looks
similar to the later part. If not, you need a longer burn-in period.
- Look at the trace plots and see if the chain is mixing. We want to
see rapid movement up and down, not slow excursions.
- See if there are any posterior correlations between parameters that
are near 1 or -1, which indicate trading off between parameters.
- Run the model with multiple chains (with different starting values)
and see that all the chains converge to the same posterior distribution.
\end{enumerate}

Let's look at some figures that illustrate these checks.

```{r mixed_diagnostics, eval=FALSE}
out.mcmc <- as.mcmc(out)[[1]]
plot(out.mcmc)
crosscorr.plot(out.mcmc)
mu1 <- c(out.mcmc[ , 1])
sig <- c(out.mcmc[ , 11])
tau <- c(out.mcmc[ , 12])
theta <- c(out.mcmc[ , 13])
plot(mu1, theta)
plot(sig, tau)
```

For the sake of time, I'll leave out the use of multiple chains, but it is 
a very good idea in general.

# HMM for pollen

In the statistics module, we worked on a HMM with a binomial model
for the data (the likelihood) and a random walk Markov model on the
logit scale as our model for the latent process. Now we're going to
make that model a bit more sophisticated.

$$ 
\mbox{logit}(\theta_t)  \sim  N(\mu + \rho (\mbox{logit}(\theta_{t-1}) - \mu), \tau^2)
$$
$$
y_{t}  \sim  \mbox{Binom}(n_{t},\theta_{t}).
$$

$\mu$ allows the model to have an arbitrary overall mean, while $\rho$ controls the auto-correlation over time.
When $\rho = 1$, we have a random walk model, and when $\rho \in (-1, 1)$ we have an autoregressive model. 

Now let's fit the
model to fake data simulated from the model (in this case we know
the statistical model is a good model for the data).

First I'll generate fake data (i.e., simulating from the forward model,
as we did yesterday).

```{r generate_DLM}
nT <- 100
n <- 50
theta0 <- 0.2
tau <- 0.2

logit <- function(x) log(x/(1-x))
expit <- function(x) exp(x) / (1+exp(x))

genDLM <- function(mu = 0, rho = 1, tau, theta0, n, nT) {
  if(length(n) == 1) n <- rep(n, nT)
  theta <- logitTheta <- y <- rep(0, nT)
  for(i in seq_len(nT)) {
    if(i == 1) {
      logitTheta[i] <- rnorm(1, mu + rho*(logit(theta0) - mu), tau)
    } else {
      logitTheta[i] <- rnorm(1, mu + rho*(logitTheta[i-1] - mu), tau)
    }
    y[i] <- rbinom(1, n[i], expit(logitTheta[i]))
  }
  return(cbind(n, y, expit(logitTheta)))
}

set.seed(0)
out <- genDLM(rho = .95, mu = -1, tau = tau, theta0 = theta0, n = n, nT = nT)
n <- out[ , 1]
y <- out[ , 2]
theta <- out[ , 3]
```

Here's the BUGS code for the model. 

```{r dlmFake}
dlmFake <- function(){
	# (hyper)priors
        mu ~ dunif(-5, 5)
        rho ~ dunif(-1, 1)
	tau ~ dunif(0, 100)
	logitTheta0 ~ dnorm(0, .000001)	
    # deterministic transformation
	tau2Inv <- 1/(tau^2)
    
	# latent process evolution and likelihood
	logitTheta[1] ~ dnorm(mu + rho*(logitTheta0 - mu), tau2Inv)
    theta[1] <- exp(logitTheta[1])/(1+exp(logitTheta[1]))  
    y[1] ~ dbin(theta[1], n[1])
	for(i in 2:nT){
		logitTheta[i] ~ dnorm(mu + rho*(logitTheta[i-1] - mu), tau2Inv)
		theta[i] <- exp(logitTheta[i])/(1+exp(logitTheta[i]))
		y[i] ~ dbin(theta[i], n[i])
	}
}
```

Now fit the model using JAGS and look at the fits.

```{r fit_dlmFake}
out <- jags(data = list(nT = nT, n = n, y = y), parameters.to.save = c('theta', 'mu', 'rho', 'tau'), 
	n.chains = 1, n.iter = 10000, n.burnin = 2000, model.file = dlmFake, DIC = FALSE)
out.mcmc <- as.mcmc(out)[[1]]
colNames <- dimnames(out.mcmc)[[2]]
whichTheta <- grep('theta', colNames)
thetaPost <- out.mcmc[ , whichTheta]
thetaNames <- dimnames(thetaPost)[[2]]
index <- gsub("theta\\[", "", thetaNames)
index <- as.numeric(gsub("\\]", "", index))
thetaPost <- thetaPost[ , order(index)]

plot(seq_len(nT), theta, type = 'l')
lines(seq_len(nT), colMeans(thetaPost), col = 'blue')
points(seq_len(nT), y/n, col = 'red')
plot(seq_len(nT), theta, type = 'l')
quants <- apply(thetaPost, 2, quantile, c(.025, .975))
# lines(seq_len(nT), quants[1, ], col = 'blue', lty = 2)
# lines(seq_len(nT), quants[2, ], col = 'blue', lty = 2)
polygon(cbind(c(seq_len(nT), nT:1, 1), c(quants[1, ], rev(quants[2, ]), quants[1, 1])), border = NA, col = 'lightblue')
lines(seq_len(nT), theta)
lines(seq_len(nT), colMeans(thetaPost), col = 'blue')
points(seq_len(nT), y/n, col = 'red')
par(mfrow = c(1, 3))
nIts <- nrow(out.mcmc)
plot(seq_len(nIts), out.mcmc[ , 1], type = 'l', main = 'mu')
plot(seq_len(nIts), out.mcmc[ , 2], type = 'l', main = 'rho')
plot(seq_len(nIts), out.mcmc[ , 3], type = 'l', main = 'tau')
print(mean(out.mcmc[201:1000, 1]))
print(mean(out.mcmc[201:1000, 2]))
print(mean(out.mcmc[201:1000, 3]))
```

Questions:
- Have we done better than the empirical estimates? Why?
- Any ideas why we might not be able to estimate $\mu$ very well?

Note: in this model, both $\rho$ and $\tau^{2}$ determine how quickly the process
changes over time. Smaller values of $\rho$ and larger values of $\tau$  allow more jumpiness while the reverse force smoother change.

Let's also see how we can choose or generate our own starting values. (See the help on `jags()` for how to set specific initial values.)

```{r own_start}
# with a function
inits <- function(){
	list('mu' = runif(1, -3, 3), 'rho' = runif(1, 0.5, 1), 'tau' = runif(1, .01, 5), 'logitTheta' = rnorm(nT, 0, 3)) 
}
# see the documentation for jags()
out <- jags(data = list(nT = nT, n = n, y = y), inits = inits, parameters.to.save = c('mu', 'tau', 'rho'), 
	n.thin = 1, n.chains = 3, n.iter = 1000, n.burnin = 0, model.file = dlmFake, DIC = FALSE)
traceplot(out, mfrow = c(1, 3), ask = FALSE)
```

Finally let's do an initial fit to real data from Tower Lake,
which we've already worked with some. First we'll read in the data.
% (formerly Aino)

```{r read_data}
data <- read.csv('data/newEngland/PollenTimeSeries.csv')
#data <- read.csv('data/upperMidwest/towerLake.csv')
names(data)[4] <- "calAge"
sub <- subset(data, sitename == 'aino')
sub$calAge <- -sub$calAge
sub <- sub[order(sub$calAge), ]
sub$n <- rowSums(sub[ , 5:14])
nT <- nrow(sub)
plot(sub$calAge, sub$oak/sub$n)
y <- sub$oak
n <- round(sub$n)
```


Here's the BUGS code for the model. 

[add mu and rho here]

```{r dlm, eval=FALSE}
dlm <- function(){
	logitTheta0 ~ dnorm(0, .000001)	
	tau ~ dunif(0, 100)
	tau2Inv <- 1/(tau^2)

	logitTheta[1] ~ dnorm(logitTheta0, tau2Inv)
    theta[1] <- exp(logitTheta[1])/(1+exp(logitTheta[1]))
    y[1] ~ dbin(theta[1], n[1])
	for(i in 2:nT){
		logitTheta[i] ~ dnorm(logitTheta[i-1], tau2Inv)
		theta[i] <- exp(logitTheta[i])/(1+exp(logitTheta[i]))
		y[i] ~ dbin(theta[i], n[i])
	}
}
```

And let's fit the model.

```{r fit_dlm, eval=FALSE}
out <- jags(data = list(nT = nT, n = n, y = y), parameters.to.save = c('tau', 'theta'), 
	n.chains = 1, n.iter = 10000, n.burnin = 2000, model.file = dlm, DIC = FALSE)
out.mcmc <- as.mcmc(out)
thetaHat <- sub$oak/sub$n
plot(sub$calAge, thetaHat, col = 'red')
lines(sub$calAge, colMeans(out.mcmc)[seq_len(nT)], col = 'blue')
quants <- apply(out.mcmc[ , seq_len(nT)], 2, quantile, c(.025, .975))
polygon(cbind(c(sub$calAge, rev(sub$calAge), sub$calAge[1]), c(quants[1, ], 
	rev(quants[2, ]), quants[1, 1])), border = NA, col = 'lightblue')
lines(sub$calAge, thetaHat, col = 'red')
lines(sub$calAge, colMeans(out.mcmc)[seq_len(nT)], col = 'blue')
title('oak at Aino Pond, MA')
abline(v = 1650 - 1950, col = 'grey')
```

# Next steps

- Challenge: How we would code up a model that treats depth
or radiocarbon years as the time index and correctly accounts for
the irregular spacing of data.
- Challenge: Let's consider some of the aspects of the data
and the scientific context that have been left out of our model for
pollen. 

