# PalEON Summer Course: R module
## August 2014
## Chris Paciorek


In this module, we'll introduce R by way of working with some pollen data.
 
First, let's make sure we have the course materials on our computers. We can download the material from Github by setting up a project in RStudio. 

Go to "File->New Project" and select "Version Control" and "Git". Enter "https://github.com/PalEON-Project/Camp2014" as the "Repository URL" and click "Create Project" (you can choose the directory in which to place the project with the "Create project as subdirectory of" option). It should create a "Camp2014" directory with all of the materials within whichever directory you chose.

Or http://www.stat.berkeley.edu/~paciorek/transfer/camp2014-statsR.zip

# Reading in data

We can read in data using the read.table() function, of which 
`read.csv()`  is a variant for reading in comma-delimited (CSV) files.
 `getwd()`  and `setwd()` help us navigate the computer's directory structure.
 `read.fwf()`  will read fixed-width format files.

```{r reading files}
getwd()

setwd('~/Desktop/camp/Camp2014/statsR')
# platform independent code:
# setwd(file.path('~', 'Desktop', 'Camp2014', 'statsR')) 

pollen <- read.csv('data/newEngland/modernPollen.csv')
ponds <- read.csv('data/newEngland/pondInfo.csv') 
#help(read.csv) 
#help(read.table) 
names(pollen) 
head(pollen) 
class(pollen)
```

# Interlude: operating at the command line


In R, as in Linux/UNIX, there are a number of tools that enhance your efficiency
:
  * The up and down arrows recall previous entries.
  * Hitting `<tab>` attempts to complete the function or object name.
  * `<Ctrl-k>` kills the remainder of the line from the cursor forward.
  * `<Ctrl-a>` goes to the beginning of the line (`<Ctrl-e>` often goes to the end but apparently not in RStudio).

More generally, typing commands and saving your operations in text files
 (aka scripts), rather than point-and-click operations, allows you to reproduce
 your work, easily redo it and modify it, and share it with others.

# R objects

Everything in R is stored as an object, each of which has a class that describes
 what the object contains and standard functions that operate on objects
 in the class.
 The basic classes are 

  * *character*  vectors: these are sets of character values (i.e., strings).
 Examples of character values are: "Sam", "0923", "Sam9", "Sam is", "Sam\t9", "Sam's the man.\nNo doubt.\n".  Each of these is a character vector of length 1.
  * *numeric* vectors (i.e., double precision real numbers)
  * *integer* vectors
  * *logical* (T/F) vectors
  * *lists*: vectors of arbitrary objects
  * *factors*: vector-like objects of categorical variables with a pre-defined set of labels
  * *functions*

More complicated objects:
  * Data frames can be thought of as a list of vectors of the same length, where the vectors may have different types.
  * Matrices are different from data frames in that all the elements are of
 a single type. Furthermore, matrices and arrays (which allow dimensions of 1, 2, 3, ....) can be thought of as vectors with information on dimensions (layout).
 If you pass a matrix into a function expecting a vector, it will just treat
 it as a vector (of concatenated columns).

We can figure out what objects exist, as well as how much memory they take
 up, and we can remove them.
```{r objects}
class(pollen) 
x <- c(7.3, 5, 11) 
class(x) 
is.data.frame(x) 
class(pollen$sitename) 
pollen$sitename 
as.character(pollen$sitename) 
class(pollen$SPRUCE) 
object.size(x) 
x <- rnorm(10000000) 
object.size(x) 
print(object.size(x), units = "Mb")  

# that code is a bit opaque and I'm lazy, so let's write our own function:
size <- function(x){   
     print(object.size(x), units = "Mb") 
} 

size(x) 
ls() 
rm(x) 
ls() 
```

# Dataframe manipulation (and some plotting)

Let's demonstrate how we can manipulate data sets.
 First, let's get some information about our two datasets and merge them
 together.

```{r dataframeInfo}
names(pollen) 
class(names(pollen)) 
names(pollen) <- tolower(names(pollen))
names(ponds) <- tolower(names(ponds)) 
cor(pollen$oak, pollen$spruce) 
dim(pollen) 
taxaCols <- 3:12
pollen$total <- rowSums(pollen[ , taxaCols]) 
pollen2 <- pollen 
pollen2[ , taxaCols] <- pollen2[ , taxaCols] / pollen2$total # careful here: I know that R will divide each numerator column by the denominator column
head(pollen2) 
full <- merge(ponds, pollen2, by.x = "site", by.y = "sitename", all.x = FALSE, all.y = TRUE) 
head(full) 
summary(full) 
taxaCols <- 9:18
cor(full[ , taxaCols])
require(fields, quietly = TRUE)
image.plot(1:10, 1:10, cor(full[ , taxaCols])) # I should change the z-scale and the color scheme
```

Now let's make some basic plots.
```{r initialPlots, fig.width=7.5}
require(maps)
plot(full$lon, full$lat, xlim = c(-73.5, -70), ylim = c(41.3,43.5), col = 'red', pch = 16) 
map('state', add = TRUE) 
par(mfrow=c(3,4)) 
for(i in taxaCols){   
  hist(full[ , i], main = names(full)[i], xlab = "proportion") 
} 
```

**Challenge**: make a single plot that shows side-by-side boxplots of the proportions for each species. Find out what input formats the `boxplot()` function can handle. Manipulate the arguments to make your figure presentable.

# More on indexing and subsetting

Now let's see some different ways to select subsets and use indexing.
 Often there are multiple ways to do things in R.
 Some are more efficient than others, which only really becomes an issue
 when working with large objects.



```{r subsetting}
sub <- full[1:5, ] 
full$lon < (-72.5) 
west <- full[full$lon < -72.5, ] 
full[full$lon < -72.5, c("oak", "pine")]
indices <- which(full$lon < -72.5)
indices
full[indices, c(15,16)]
west2 <- subset(full, lon < -72.5, select = c("oak", "pine"))
```

**Challenge**: extract the even rows of the full dataset. Consider the 
`rep()` and `seq()` functions.

R does not provide a built-in function for sorting data frames.
```{r sorting}
full <- full[order(full$lat, full$lon), ]
head(full)[ , 1:8]
```

That's a bit frustrating but easily fixable once we talk more about writing
 functions. (Also see `arrange()` in the `plyr` package.)

# Vectorized operations and the apply function

Many R functions operate on vectors without having to loop over the indices,
 and in fact vectorized operations are generally much faster than using
 a loop.

```{r vectorized operations}
log(full$oak)
full$coldSpp <- full$beech + full$hemlock + full$spruce 
full$west <- full$lon < -72.5 
full$north <- full$lat > 42.5 
full$north <- FALSE # example of recyling 
full$north[full$lat > 42.5] <- TRUE  
full$nw <- full$west & full$north 
round(full$lat, digits = 1) 
```

Useful commands are `with()` and `within()`, which allow you to avoid a bunch of typing. The basic idea is to do a bunch of stuff all within a certain context.

```{r with}
full <- within(full, warmSpp <- oak + hickory)
full$warmSpp[1:8]
```

R has a feature called recycling by which it will fill in additional entries in vectors, matrices, data frames:

```{r recycling, fig.width=4}
mat <- matrix(1:2, nrow = 4, ncol = 4) # what do you think will happen?
mat <- matrix(1:3, nrow = 4, ncol = 4) 
```

One useful type of data structure in R is a list.
 A list is a vector of potentially unrelated items.
 Lists are commonly used as the output from functions.

```{r lists}
myList <- list(a = 7, b = c(8, 9, 11), d = "wtfit?", e = list(first = 7,
 second = 8)) 
myList$b 
myList[[2]] 
myList$e$first
is.list(full)
myModel = lm(spruce ~ lat, data = full) 
is.list(myModel) 
names(myModel) 
myModel$coefficients 
```

We can use the apply function to "lapply" an arbitrary function to each element of the list. This is often faster (and is easier to write) than using a loop.

```{r apply}
myList = list(1:3, 4:8, 101, 111:150) 
lapply(myList, max) # vs.
out = rep(NA, length(myList)) 
for(i in 1:length(myList))   
   out[i] <- max(myList[[i]]) 
```

**Challenge**: how can I find the median abundance of each species across all the ponds?
 
# Functions

We've already seen a bit of functions: we wrote our own little function
 and we've used lots of functions from the core R software and additional
 packages.

Functions allow one to modularize their code and are at the core of good
 programming.
 
Let's create a function to plot the proportion of a given taxon on a map. Once we we've done this, we can use the function on each taxon to create one map per taxon.

First, we'll prototype the code for a single taxon.

What do we need to decide in terms of what our function is going to do, what its arguments should be, and what steps the function should carry out? 


```{r function_prototype}
require(maps)
x <- full$lon
y <- full$lat
z <- full$oak
propBreaks = c(0, 0.01, 0.05, 0.10, 0.15, 0.2, 0.3, 0.4, 0.5, 0.6, 0.8, 1)
colors <- rev(terrain.colors(length(propBreaks)))
cats <- cut(z, breaks = propBreaks)
cats <- as.integer(cats)
plot(x, y, xlim = c(-73.5, -70), ylim = c(41.3, 43.5), col = colors[cats], pch = 16)
map("state", add = TRUE)

```

Now let's work through how we would make a function out of this.

```{r full_function}

```

# Plotting bells and whistles

Let's take our previous plots and gussy them up for the ball, or at least fit for a trip into the village.

Note that more recent R packages provide more visually-appealing graphics and more elegant syntax, in particular the `ggplot2` package. We'll likely see more of that later this week.

```{r advanced plots}
pdf('treeComp.pdf', height = 4, width = 7)
par(mfrow = c(2,5), mai = c(.25,.25,.3,.1), omi = c(.3, .3, .4, 0))
for(i in taxaCols)
      hist(full[ , i], xlim = c(0, 1), xlab = '', ylab = '', main = names(full)[i])
title(ylab = 'frequency', outer = TRUE, line = 1, cex.lab = 1.5)
title(xlab = 'proportion of pollen', outer = TRUE, line = 1, cex.lab = 1.5)
title(main = paste('Pollen composition of ', nrow(full), ' ponds in New England', sep = ''), 
     outer = TRUE, line = 1, cex.main = 1.5)
dev.off()
#help(par)
par()$mfrow
```

**On your own**: Take a look at the sort of graphics that R can create by invoking `demo(graphics)`, `demo(persp)`, `demo(image)`.

# Packages and loading data from other software

As part of preparing for the course, you've probably already downloaded
 and installed an R package from the R archive, CRAN.
 Packages provide additional functionality and are the "killer app"
 of R (think the App Store for Apple).
 The packages are contributed by users and R developers and vary in quality.
 They implement most statistical algorithms (and other algorithms and utilities)
 that now exist.
 So if you're about to code something, it's worth checking to see if it's
 already been done.
 And if you develop or implement a new method, consider contributing it
 to CRAN as a package.

A useful place to look for answers to questions, find out about functionality
 and pose questions is the [R mailing list and its archive](http://tolstoy.newcastle.edu.au/R/).  If you do post, follow the guidelines and provide enough information about the issue for those reading the list to be able to understand the question  and reproduce any problems.
 
One very useful package is the `foreign` package, which contains interfaces that can read and write data formats for SAS, Stata, DBF, etc.
 For those who use netCDF and HDF files, there are packages that deal with
 those too (*ncdf4* in the case of netCDF).

```{r packages}
library(foreign) 
#library(help = foreign) 
```

# Finishing up

You can save your code in text files and read it all in at once, using `source()`: e.g.,
```{r source}
source('silly.R')
```

You can save some or all of your objects in a compressed R data file
 and load the objects back into R on any operating system, regardless of the system used to create the file.

```{r save}
save(full, pollen, ponds, file = 'pollenAnalData.Rda') 
# save.image('pollenFullAnal.Rda')
rm(list = ls()) 
ls() 
load('pollenAnalData.Rda') 
ls() 
```

You can write out CSV and other delimited files with `write.table()` and `write.csv()`.

